---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 6 @ 11:59PM
author: Seamus Gallivan
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the tidyverse.

**Solution**

```{r}
system.time(read.csv("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
system.time(read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
system.time(fread("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
```

fread is the fastest at 0.871 seconds. read_csv is not far behind at 2.184 seconds. The base R read.csv is significantly slower, taking almost 42 seconds. 

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 

**Solution**

```{r}
icustays_tble <- read_csv("/mnt/mimiciv/1.0/icu/icustays.csv.gz")
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 

**Solution**

53,150 unique 'subject_id'. Yes, from quick glimpse at the tibble below,
subject_id number 10002428 has 4 ICU stays.

```{r}
count(icustays_tble, subject_id)
```

3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.

**Solution**

```{r}
icustays_tble <- icustays_tble %>%
  arrange(intime) %>%
  distinct(subject_id, .keep_all = TRUE)
```


## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.

**Solution** 

```{r}
admissions_tble1 <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadm_id`.

```{r}
admissions_tble <- semi_join(admissions_tble1, icustays_tble, 
                              by = c("subject_id", "hadm_id"))
admissions_tble %>% arrange(subject_id)
```


3. Summarize the following variables by graphics. 

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  

**Solution** 

Admission month, monthday and weekly are all uniformly distributed. Interestingly enough, admission hour peaks at midnight, decreases until it spikes at 7am, jumps back down at 8am and gradually increases until 3pm where it stays constant until midnight.   

```{r}
#admission year  
c <- ggplot(admissions_tble, aes(year(admittime)))
c + geom_bar()

#admission month
d <- ggplot(admissions_tble, aes(lubridate::month(admittime, label = T)))
d + geom_bar() + labs(x = "Month")

#admission month day 
a <- ggplot(admissions_tble, aes(lubridate::mday(admittime)))
a + geom_bar() + labs(x = "month day")

#admission week day 
b <- ggplot(admissions_tble, aes(lubridate::wday(admittime, label = T)))
b + geom_bar() + labs(x = "weekday")

#admission hour
e <- ggplot(admissions_tble, aes(lubridate::hour(admittime)))
e + geom_bar() + labs(x = "hour")

```

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

**Solution**

```{r}
patients_tble1 <- read_csv("/mnt/mimiciv/1.0/core/patients.csv.gz") 
patients_tble <- semi_join(patients_tble1, icustays_tble, by = "subject_id")
patients_tble %>% arrange(subject_id)
```

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

**Solution**

```{r}
#summarize gender 
gender <- ggplot(patients_tble, aes(gender))
gender + geom_bar()

#summarize anchor_age
age <- ggplot(patients_tble, aes(anchor_age))
age + geom_histogram()
```


## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

**Solution**

1630 rows (excluding the header)

```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | tail -n +2 | wc -l"
    ), 
  intern = TRUE
)
```

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.
    
**Solution**

```{r}
labevents_tble1 <- 
  read_csv(str_c(mimic_path, "/hosp/labevents_filtered_itemid.csv.gz"),
      col_select = c(subject_id, itemid, charttime, valuenum),
      col_types = cols_only(subject_id = col_double(), 
                            itemid = col_double(), 
                            charttime = col_datetime(), 
                            valuenum = col_double()),
      lazy = TRUE) %>%
  semi_join(icustays_tble, by = c("subject_id")) 
```

```{r}
d_labitems <- read_csv(str_c(mimic_path, "/hosp/d_labitems.csv.gz"))
labevents_tble1 <- left_join(labevents_tble1, d_labitems, by = "itemid")
```

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 

```{r}
labevents_tble <- labevents_tble1 %>%
  # pull in intime from icu stay
  left_join(select(icustays_tble, subject_id, intime), by = "subject_id") %>%
  #only lab measures after intime
  filter(charttime >= intime) %>%
  #first lab measurement after intime for each subjectid x itemid combo
  group_by(subject_id, itemid) %>%
  arrange(charttime, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup() %>% 
  select(subject_id, label, valuenum) %>% 
  pivot_wider(names_from = label, values_from = valuenum) 
labevents_tble

```

4. Summarize the lab measurements by appropriate numerics and graphics. 

```{r}
lab_box <- labevents_tble %>% 
  select(2:11) %>% 
  pivot_longer(1:10, names_to = "label", values_to = "values")

fivefour <- ggplot(lab_box, aes(x = label, y = values, color = label)) +
  geom_boxplot()
fivefour + coord_flip() + xlab("Lab Measurement") +
  ylab("Results (mixed units)") + ggtitle("Lab Measurements Boxplots")
```

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

**Solution**

```{r}
d_items <- read_csv(str_c(mimic_path, "/icu/d_items.csv.gz"))

chartevents_tble1 <- 
  read_csv(str_c(mimic_path, "/icu/chartevents_filtered_itemid.csv.gz")) %>%
  left_join(d_items, by = "itemid") %>%
  semi_join(icustays_tble, by = "subject_id")
```

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay.

```{r}
chartevents_tble <- chartevents_tble1 %>%
  # pull in intime from icu stay
  left_join(select(icustays_tble, subject_id, intime), by = "subject_id") %>%
  #only lab measures after intime
  filter(charttime >= intime) %>%
  #first lab measurement after intime for each subjectid x itemid combo
  group_by(subject_id, itemid) %>%
  arrange(charttime, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup() %>% 
  select(subject_id, valuenum, label) %>% 
  pivot_wider(names_from = label, values_from = valuenum)
chartevents_tble
```

3. Summarize these vital measurements by appropriate numerics and graphics. 

```{r}
chart_box <- chartevents_tble %>% 
  select(2:6) %>% 
  pivot_longer(1:5, names_to = "label", values_to = "values")
sixthree <- ggplot(chart_box, aes(x = label, y = values, color = label)) +
  geom_boxplot()
sixthree + coord_flip() + xlab("Vital Measurement") +
  ylab("Results (mixed units)") + 
  ggtitle("Vital Measurements Boxplots (outliers excluded") +
  ylim(0, 300)
```

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

**Solution**

```{r}
mimic_icu_cohort <- icustays_tble %>%
  left_join(patients_tble, by = "subject_id") %>%
  left_join(admissions_tble, by = c("hadm_id", "subject_id")) %>% 
  left_join(labevents_tble, by = "subject_id") %>% 
  left_join(chartevents_tble, by = "subject_id") %>% 
  mutate(age_hadm = anchor_age + year(admittime) - anchor_year) %>% 
  filter(age_hadm > 18) %>% 
  mutate(thirty_day_mort = ifelse(is.na(deathtime), Inf, ifelse(as.Date(deathtime) -
           as.Date(admittime) <= 30, "Yes", "No")))
mimic_icu_cohort
```


## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

**Solution**

Ethnicity
```{r}
mort_eth <- mimic_icu_cohort %>% 
  select(ethnicity, thirty_day_mort) %>% 
  ggplot(aes(thirty_day_mort))
mort_eth + geom_bar(aes(fill = ethnicity)) + 
  scale_fill_discrete(name="Ethnicity") + xlab("Thirty day mortality")
```

Language
```{r}
mort_lang <- mimic_icu_cohort %>% 
  select(language, thirty_day_mort) %>% 
  ggplot(aes(thirty_day_mort))
mort_lang + geom_bar(aes(fill = language)) + 
  scale_fill_discrete(name="Language") + xlab("Thirty day mortality")
```

Insurance
```{r}
mort_ins <- mimic_icu_cohort %>% 
  select(insurance, thirty_day_mort) %>% 
  ggplot(aes(thirty_day_mort))
mort_ins + geom_bar(aes(fill = insurance)) + 
  scale_fill_discrete(name="Insurance") + xlab("Thirty day mortality")
```

Marital status
```{r}
mort_mar <- mimic_icu_cohort %>% 
  select(marital_status, thirty_day_mort) %>% 
  ggplot(aes(thirty_day_mort))
mort_mar + geom_bar(aes(fill = marital_status)) + 
  scale_fill_discrete(name="Marital Status") + xlab("Thirty day mortality")
```


- `thirty_day_mort` vs first lab measurements

**Solution**

```{r}
mort_lab <- mimic_icu_cohort %>% 
  select(27:36, thirty_day_mort) %>% 
  pivot_longer(1:10, names_to = "measurement", values_to = "value") %>% 
  ggplot(aes(thirty_day_mort))
mort_lab + geom_bar(aes(fill = measurement)) + 
  scale_fill_discrete(name="Measurement label") + xlab("Thirty day mortality")
  
```

Gender
```{r}
mort_gen <- mimic_icu_cohort %>% 
  select(gender, thirty_day_mort) %>% 
  ggplot(aes(thirty_day_mort))
mort_gen + geom_bar(aes(fill = gender)) + 
  scale_fill_discrete(name="Gender") + xlab("Thirty day mortality")
```

Age at hospital admission 
```{r}
mort_age <- mimic_icu_cohort %>% 
  select(age_hadm, thirty_day_mort) %>% 
  ggplot(aes(age_hadm))
mort_age + geom_histogram(aes(fill = thirty_day_mort)) + 
  scale_fill_discrete(name="Age at admission") + xlab("Thirty day mortality")
```


- `thirty_day_mort` vs first vital measurements

**Solution**

```{r}
mort_vital <- mimic_icu_cohort %>% 
  select(37:41, thirty_day_mort) %>% 
  pivot_longer(1:5, names_to = "measurement", values_to = "value") %>% 
  ggplot(aes(thirty_day_mort))
mort_vital + geom_bar(aes(fill = measurement)) + xlab("Thirty day mortality") +
  scale_fill_discrete(name="Measurement type")
```

- `thirty_day_mort` vs first ICU unit

**Solution**

```{r}
mort_unit <- mimic_icu_cohort %>% 
  select(first_careunit, thirty_day_mort) %>% 
  ggplot(aes(thirty_day_mort))
mort_unit + geom_bar(aes(fill = first_careunit)) + 
  xlab("Thirty day mortality") + scale_fill_discrete(name="Unit type")
```

